\section{Evaluation}
\label{sec:evaluation}

\subsection{Models compared}

Four configurations are evaluated. They span two architectural choices --- a small custom
transformer versus a much larger pretrained model --- and, within each choice, a comparison of
training from scratch (or zero-shot) against a model that has been adapted to the task data.

\begin{table}[h]
\centering
\small
\begin{tabular}{p{3.4cm}p{3.0cm}p{2.6cm}p{4.2cm}}
\toprule
\textbf{Model} & \textbf{Initialisation} & \textbf{Training data} & \textbf{Role} \\
\midrule
Custom transformer (scratch) & Random ($\mathcal{N}(0, 0.02)$) & Synthetic + public QA &
Fulfils mandatory ``train one model from scratch'' criterion. \\
Custom transformer (GPT-2 init) & GPT-2 small weights & Synthetic + public QA &
Same architecture, warm-started from a pretrained checkpoint. \\
Qwen2.5-3B-Instruct & Pretrained, frozen & --- & Strong zero-shot reference. \\
Qwen2.5-3B-Instruct + LoRA & Pretrained + LoRA adapters & Synthetic & Parameter-efficient
fine-tune on the synthetic task data \cite{hu2022lora}. \\
\bottomrule
\end{tabular}
\caption{The four configurations under comparison. Only the first two are decoder-only models
trained on this project's hardware; the second pair establishes how close a small custom model
gets to a much larger off-the-shelf instruction-tuned model.}
\label{tab:models}
\end{table}

The two custom configurations isolate the contribution of pretraining on a generic web
corpus, holding the architecture constant. The two Qwen configurations isolate the
contribution of LoRA fine-tuning on the synthetic data, holding the base model constant. All
four models are wrapped behind the same RAG retriever, so any difference reflects generation
quality rather than retrieval.

\subsection{Metrics}

Four complementary metrics are reported. Two are intrinsic to the language model
(perplexity, format compliance), and two are extrinsic, judging the usefulness of the output
to a student (correctness, retrieval relevance).

\paragraph{Validation perplexity.}
Standard intrinsic metric, computed on the held-out split. It reflects how well the model
predicts the next token but, on its own, has known weaknesses as a proxy for output
quality~\cite{zheng2023judging}; here it is reported mainly as a sanity check that training
converged and as a means of comparing the two custom variants.

\paragraph{Format compliance.}
The fraction of generated outputs that parse as valid for their declared task type. For
\emph{quiz}, the post-prompt segment must be a JSON object with the required keys and four
distinct options. For \emph{review}, the output must contain a numeric score and a
non-empty review text. For \emph{flashcard}, the output must contain both a front and a back
delimited as in the training data. \emph{Explanation} is excluded since it has no rigid
schema. This metric directly measures whether the model has learnt the task structure rather
than merely fluent prose.

\paragraph{Answer correctness (LLM-as-judge).}
For each test query, the model's answer and a held-out reference answer are presented to
GPT-4 with a scoring rubric that returns an integer from 1 (incorrect) to 5 (fully correct
and well-explained). LLM-as-judge has been shown to correlate well with human ratings on
generative tasks~\cite{zheng2023judging}. The same prompt is used for all four models so
that the absolute scores are comparable.

\paragraph{Retrieval relevance.}
Independent of the generator, the retrieval component is evaluated using normalised
discounted cumulative gain at $k=3$ (NDCG@3) against a manually annotated relevance set of
30 course-specific queries. This metric is shared across all four models because they use
the same retriever.

\subsection{Results}

\begin{table}[h]
\centering
\small
\begin{tabular}{lcccc}
\toprule
\textbf{Model} & \textbf{Val.\ ppl} ($\downarrow$) & \textbf{Format compl.} ($\uparrow$) &
\textbf{Judge score} ($\uparrow$) & \textbf{Inference (s/req)} \\
\midrule
Custom (scratch)            & \textit{tbd} & \textit{tbd} & \textit{tbd} & \textit{tbd} \\
Custom (GPT-2 init)         & \textit{tbd} & \textit{tbd} & \textit{tbd} & \textit{tbd} \\
Qwen2.5-3B-Instruct         & ---          & \textit{tbd} & \textit{tbd} & \textit{tbd} \\
Qwen2.5-3B-Instruct + LoRA  & \textit{tbd} & \textit{tbd} & \textit{tbd} & \textit{tbd} \\
\bottomrule
\end{tabular}
\caption{Comparison of the four configurations. Validation perplexity is not reported for the
zero-shot Qwen baseline since it is not trained on the task data.}
\label{tab:results}
\end{table}

\begin{table}[h]
\centering
\small
\begin{tabular}{lcc}
\toprule
\textbf{Retriever} & \textbf{NDCG@3} ($\uparrow$) & \textbf{Recall@5} ($\uparrow$) \\
\midrule
\texttt{BAAI/bge-base-en-v1.5} (used)  & \textit{tbd} & \textit{tbd} \\
\texttt{all-MiniLM-L6-v2} (CPU baseline) & \textit{tbd} & \textit{tbd} \\
TF--IDF                                 & \textit{tbd} & \textit{tbd} \\
\bottomrule
\end{tabular}
\caption{Retrieval-only ablation on 30 manually annotated queries. The TF--IDF row tests
whether the dense retriever adds value over a classical sparse baseline.}
\label{tab:retrieval}
\end{table}

\textit{[Final numbers will be filled in once all four training runs and the LoRA adaptation
have finished. The placeholder rows indicate the structure of the comparison.]}

\subsection{Discussion}

The expected ordering of the four models on the judge score is:
Qwen+LoRA $\geq$ Qwen zero-shot $>$ Custom (GPT-2 init) $>$ Custom (scratch). The custom
models are not expected to outperform the Qwen baseline on absolute correctness; the
$\sim$26\,M-parameter model has roughly two orders of magnitude fewer parameters than the
3\,B Qwen baseline, and a fair fraction of its capacity is consumed by simply learning
fluent English. The interesting comparisons are therefore the two pairwise differences:
how much warm-starting from GPT-2 buys over random initialisation, and how much LoRA
fine-tuning buys over the zero-shot Qwen baseline.

Format compliance is expected to favour the custom models more than the judge score does:
the synthetic training corpus exposes the custom models to thousands of valid quiz JSON
objects, while the zero-shot Qwen has only the few-shot examples in the prompt. This is the
practical justification for the small model in deployment. Even if its prose is less polished,
a model that always emits parseable JSON is more useful inside the quiz tab than a model
whose output frequently breaks the parser.

Validation perplexity is reported as a training diagnostic only. During development, the
format-compliance percentage on the validation set was found to be a much better
checkpoint-selection signal than perplexity --- runs with similar perplexity differed by
tens of percentage points in JSON validity --- and was used as the primary stopping
criterion.
