\section{Model Implementation}
\label{sec:model}

\subsection{Architecture}

The generator is a decoder-only transformer~\cite{vaswani2017attention}. The structural
backbone --- causal self-attention, pre-norm LayerNorm, GELU feed-forward blocks, weight tying
between the input embedding and the language model head, and the $\mathcal{N}(0, 0.02)$
initialisation scheme with the residual-scaled projection variance \cite{radford2019gpt2}
and the \texttt{c\_attn}/\texttt{c\_proj} naming convention --- follows
nanoGPT~\cite{karpathy2022nanogpt}, a clean reimplementation of the GPT-2 small design.
Two departures from that baseline are used in the reported runs: rotary positional
embeddings, which replace GPT-2's learned absolute positions, and a key/value cache for
faster autoregressive decoding. The codebase additionally supports RMSNorm, a SwiGLU
feed-forward variant, and stochastic depth, all introduced in modern open-weight models
such as LLaMA~\cite{touvron2023llama}; these are available behind flags but were left
disabled in the reported runs so that the warm-started checkpoint stays binary-compatible
with GPT-2's pretrained weights. The forward pass is sketched in
Figure~\ref{fig:architecture}.

\begin{figure}[h]
\centering
\begin{tikzpicture}[node distance=0.4cm]

    % Input
    \node[data] (tokens) {Token IDs \\ \footnotesize $(B, T)$};
    \node[block, above=of tokens] (embed) {Token embedding \\ \footnotesize tied to LM head};

    % Block (single)
    \node[draw=black!50, dashed, rounded corners=4pt, fill=gray!3,
          minimum width=8.5cm, minimum height=5.6cm,
          above=of embed, yshift=0.3cm] (blockbox) {};
    \node[smallnote, anchor=south east] at ($(blockbox.south east)+(-0.2,0.15)$) {$\times N$ blocks};

    \node[block, above=1.4cm of embed, fill=blue!4] (norm1) {LayerNorm};
    \node[block, above=of norm1, fill=blue!8] (attn) {Causal multi-head attention \\ \footnotesize with RoPE on Q,K (+ KV cache)};
    \node[block, above=of attn, fill=blue!4] (norm2) {LayerNorm};
    \node[block, above=of norm2, fill=blue!8] (ffn) {GELU FFN \\ \footnotesize hidden $4d$};

    % Final norm + head
    \node[block, above=1.6cm of ffn] (normf) {LayerNorm};
    \node[block, above=of normf] (head) {LM head \\ \footnotesize logits $(B, T, V)$};

    % Edges
    \draw[arrow] (tokens) -- (embed);
    \draw[arrow] (embed) -- (norm1);
    \draw[arrow] (norm1) -- (attn);
    \draw[arrow] (attn) -- (norm2);
    \draw[arrow] (norm2) -- (ffn);
    \draw[arrow] (ffn) -- (normf);
    \draw[arrow] (normf) -- (head);

\end{tikzpicture}
\caption{Forward pass of the custom transformer. Pre-norm LayerNorm wraps causal multi-head
attention (with RoPE applied to queries and keys) and a GELU feed-forward block. $V = 50\,261$
tokens, $d = 768$, $N = 12$.}
\label{fig:architecture}
\end{figure}

\paragraph{Rotary positional embeddings.}
Position information is injected by rotating the query and key projections by an angle that
depends on the absolute position of each token~\cite{su2024roformer}. Unlike learned absolute
embeddings, RoPE makes the relative position between two tokens an explicit feature of the
attention dot product, which extrapolates better to sequence lengths not seen during training
and is the dominant choice in current open-weight models.

\paragraph{Available architecture variants.}
Three modern alternatives are implemented and toggleable at construction time but are not
enabled in the reported runs. RMSNorm~\cite{zhang2019rmsnorm} drops the mean-centring term
of LayerNorm; the SwiGLU gating variant~\cite{shazeer2020glu}, $\mathrm{FFN}(x) = (x W_1
\odot \mathrm{SiLU}(x W_2)) W_3$ with inner dimension $\approx \tfrac{8}{3}d$, replaces the
GELU FFN; and stochastic depth~\cite{huang2016stochastic} drops residual sub-blocks with a
linearly increasing probability across depth. These are all enabled by command-line flags
(\texttt{--use-rmsnorm}, \texttt{--ffn-variant swiglu}, \texttt{--stochastic-depth}); leaving
them off keeps the architecture binary-compatible with GPT-2 small, which is required for
the warm-start checkpoint to load cleanly.

\paragraph{KV cache.}
At inference time, autoregressive decoding recomputes the same key and value projections for
every previously generated token. The implementation caches them so each new step only
projects the most recent token, reducing the per-step cost from $O(T^2)$ to $O(T)$ when
sampling long outputs (notably for the \emph{review} task).

\paragraph{Hyperparameters.}
Table~\ref{tab:hyperparams} summarises the architectural settings. The model has roughly
124\,M parameters with weight tying, matching the GPT-2 small geometry.

\begin{table}[h]
\centering
\begin{tabular}{ll}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Layers ($N$)                & 12 \\
Attention heads             & 12 \\
Embedding dimension ($d$)   & 768 \\
FFN hidden dim              & 3\,072 ($4d$, GELU) \\
Context length              & 512 tokens \\
Positional encoding         & RoPE \\
Normalisation               & LayerNorm (pre-norm) \\
Linear biases               & enabled (GPT-2 compatibility) \\
Vocabulary                  & GPT-2 BPE + 4 task tokens (50\,261) \\
\bottomrule
\end{tabular}
\caption{Architecture settings of the custom transformer (GPT-2 small geometry).}
\label{tab:hyperparams}
\end{table}

Attention is computed via \texttt{torch.nn.functional.scaled\_dot\_product\_attention}, which
dispatches to a FlashAttention kernel when one is available and falls back to a memory-safe
reference implementation otherwise.

\subsection{Tokenisation and task tokens}
\label{sec:tokenizer}

The four task formats --- explanation, quiz, review, flashcard --- are signalled to the model
by four dedicated special tokens that prefix every example. Reusing GPT-2's existing BPE
vocabulary avoided retraining a tokeniser from scratch, but four IDs had to be appended to
the vocabulary:

\begin{center}\small
\begin{tabular}{lll}
\toprule
\texttt{<|explain|>} & ID 50\,257 & free-form explanation \\
\texttt{<|quiz|>}    & ID 50\,258 & multiple-choice quiz \\
\texttt{<|review|>}  & ID 50\,259 & long-answer + scored review \\
\texttt{<|flashcard|>} & ID 50\,260 & front/back card \\
\bottomrule
\end{tabular}
\end{center}

The implementation rebuilds the tiktoken \texttt{gpt2} encoding with the four extra entries
in the \texttt{special\_tokens} table, preserving the original 50\,256 BPE merges and
the \texttt{<|endoftext|>} marker (ID 50\,256) used as end-of-sequence. The model's embedding
table and language model head are sized to 50\,261 to match. Conditioning on a task token in
this way is a lightweight alternative to training separate heads per task: a single forward
pass covers all four activities, and the routing decision is made by one BPE token at the
start of each prompt rather than by a learned classifier head.

\subsection{Synthetic training data}
\label{sec:training_data}

A model trained from random initialisation on the public QA corpora alone would learn to
answer general machine-learning questions, but it would not learn to emit the four task
formats above. To produce supervision for those formats, an instruction-tuned local model,
\texttt{Qwen2.5-7B-Instruct}~\cite{yang2024qwen25}, was used to generate task-specific
training examples from the chunks indexed in Section~\ref{sec:data}. Running the data
generator locally avoided API costs and let the same chunk be expanded into all four task
types in a single pass over the corpus. Figure~\ref{fig:datagen} sketches the pipeline.

\begin{figure}[h]
\centering
\begin{tikzpicture}[node distance=0.5cm and 0.7cm]
    \node[data] (chunks) {Indexed chunks \\ \footnotesize \texttt{chunks.json}};
    \node[block, right=of chunks] (qwen) {Qwen2.5-7B \\ \footnotesize 4 task prompts};

    \node[data, right=1.3cm of qwen, yshift=1.5cm]   (e) {\texttt{explanation}};
    \node[data, right=1.3cm of qwen, yshift=0.5cm]   (q) {\texttt{quiz}};
    \node[data, right=1.3cm of qwen, yshift=-0.5cm]  (r) {\texttt{review}};
    \node[data, right=1.3cm of qwen, yshift=-1.5cm]  (f) {\texttt{flashcard}};

    \node[data, right=1.5cm of q, yshift=-0.5cm] (jsonl) {\texttt{train.jsonl}};

    \draw[arrow] (chunks) -- (qwen);
    \draw[arrow] (qwen.east) -- (e.west);
    \draw[arrow] (qwen.east) -- (q.west);
    \draw[arrow] (qwen.east) -- (r.west);
    \draw[arrow] (qwen.east) -- (f.west);
    \draw[arrow] (e.east) -- (jsonl.north west);
    \draw[arrow] (q.east) -- (jsonl.west);
    \draw[arrow] (r.east) -- (jsonl.west);
    \draw[arrow] (f.east) -- (jsonl.south west);
\end{tikzpicture}
\caption{Synthetic data generation. Each indexed chunk yields one example per task type.
Resume state is tracked by SHA-1 hash of the chunk text, so an interrupted run continues at
chunk granularity rather than chapter granularity.}
\label{fig:datagen}
\end{figure}

A single line of \texttt{train.jsonl} stores the structured task fields together with a raw
\texttt{"text"} field that captures Qwen's generation. At training time
\texttt{dataset.py::format\_for\_training} re-renders each example into the canonical string
that the model actually sees; for a \emph{quiz} record this string is

\begin{quote}\footnotesize\ttfamily
<|quiz|> Question: $\langle q\rangle$ \\
A) $\langle$option 0$\rangle$ \\
B) $\langle$option 1$\rangle$ \\
C) $\langle$option 2$\rangle$ \\
D) $\langle$option 3$\rangle$ \\
Answer: $\langle$letter$\rangle$) $\langle$correct option$\rangle$. $\langle$explanation$\rangle$<|endoftext|>
\end{quote}

(\emph{explain}, \emph{review} and \emph{flashcard} use analogous templates with a different
prefix token.) Re-rendering at load time means the structured fields stay authoritative ---
format-validity metrics (Section~\ref{sec:evaluation}) read them directly rather than
re-parsing the generated string. The \texttt{chunk\_hash} provides a stable resume key:
an earlier implementation that tracked progress by chapter source path skipped every chunk
in a chapter once the first one was processed; switching to a SHA-1 hash of the chunk text
fixed this and made interrupted runs safely restartable.

For the explanation and flashcard tasks, Qwen is asked to write a request and the
corresponding response in the voice of a teaching assistant. For \emph{quiz}, the prompt
constrains the output to a strict JSON schema with four options and a \texttt{correct\_index}
field. For \emph{review}, Qwen first generates a question and a plausible (and deliberately
imperfect) student answer, then writes a graded review of that answer with a numeric score
and a reference answer. The reference answer is what the model is later judged against.

The validity of the generated data is checked by a \texttt{--validate} pass that re-parses
each line and rejects entries with malformed task fields, missing context, or a length below
a minimum threshold; rejected entries are not used in training.

\subsection{Training procedure}

The model is trained with the standard autoregressive cross-entropy objective over the
re-rendered \texttt{text} field. Loss is computed on every token (no prefix masking), since
the task prefix and context are themselves part of the in-distribution sequences the model
needs to learn to recognise. Label smoothing of $0.05$ is applied to discourage extreme
logit values and improve calibration on the structured tasks where an overconfident wrong
token breaks the expected output shape.

Two custom-transformer runs are reported. The \texttt{scratch\_125m} run trains the model
from random $\mathcal{N}(0, 0.02)$ initialisation. The \texttt{gpt2\_warm\_full} run starts
from the same architecture but loaded with weights from \texttt{openai-community/gpt2}; the
warm start is the reason the architecture is kept binary-compatible with GPT-2 small.

\paragraph{Optimisation.}
The optimiser is AdamW~\cite{loshchilov2019decoupled} with $\beta_1 = 0.9$, $\beta_2 = 0.95$
and weight decay $0.1$. Parameter groups are split by parameter \emph{name}, not by tensor
shape: every embedding-like parameter is excluded from weight decay regardless of its
dimensionality. This is a small but important correction relative to the common shape-based
heuristic, which silently regularises the embedding table on tied-weight models. Gradient
norms are clipped to $1.0$. The \texttt{scratch\_125m} run uses a peak learning rate of
$3 \times 10^{-4}$ with 200 linear warmup steps; \texttt{gpt2\_warm\_full} drops the peak to
$1 \times 10^{-4}$ with 100 warmup steps, since the GPT-2 weights already sit close to a
useful basin and a smaller step keeps the warm-started transformer from diverging early.
Both runs share a cosine decay to $\mathrm{min\_lr} = 3 \times 10^{-5}$ over a total budget
of 8\,000 optimiser steps.

\paragraph{Hardware utilisation.}
Training runs on a single NVIDIA RTX 3090. Forward and backward passes use \texttt{bfloat16}
autocasting; the master weights stay in \texttt{float32}. The model is wrapped in
\texttt{torch.compile} to fuse the attention and FFN kernels. With a per-step batch size of
16 and gradient accumulation of 4, the effective batch size is 64 sequences of 512 tokens.

\paragraph{Training-time RAG context mixing.}
Half of the training prompts (\texttt{--rag-context-prob 0.5}) are prefixed with the chunk
that originally seeded the synthetic example, formatted as a \texttt{Context: $\langle$chunk$\rangle$}
block. The other half are presented bare, exactly as the model will see them when retrieval
is unavailable at inference time. Mixing both regimes during training stops the model from
overfitting to the always-context distribution and was found, in earlier exploratory runs,
to make the warm-started checkpoint robust to with/without-context prompts at evaluation
time without changing the model architecture.

\paragraph{Stabilisation tricks.}
An exponential moving average (EMA) of the parameters is maintained alongside the live
weights, with a decay of $0.999$. The EMA copy is recorded for validation and is used as
the source of inference weights when its loss tracks the live weights. Early stopping on
validation loss has a patience of 5 validation intervals (one validation interval = 400
optimiser steps), which prevents long-tail overfitting on the smaller in-domain split.

\paragraph{Logging.}
Training and validation curves are streamed to Weights \& Biases. In addition to standard
loss and perplexity, the validation loop logs per-task validation losses (one per task
prefix) and a small number of decoded samples per task, so that drift in any single task
format is visible in the dashboard rather than hidden inside an aggregate metric.
